A pipeline to analyse time-course gene expression data

### 2019-05-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@

The per-time point analysis is often the entire approach advocated for dealing with small time-course datasets, where there are only a few time points \textbf{FIXME add citation to edgeR/limma vignette or DESEQ2}. For long time-course datasets, however, such an approach results in creating many different tests, for example for every pair of adjacent timepoints, the results of which are difficult to integrate. We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.

@@

\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of these types of approaches. In both situations, we first set up a object (a \texttt{splines\_model} object) to hold the meta data, as well as information for fitting the spline model.

@@

@@

\texttt{moanin} provides a simple interface to perform a time by time differential
expression analysis. Under the hood, it simply calls \texttt{limma} \citep{ritchie:limma} on the set of
contrasts provided. Here, we show an example between the control mouse and the
mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
(H5N1), but the function works with any form contrasts.

@@

The distribution of number of genes found differentially expressed by
considering each time-point independantly highlights the challenges of such
approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
genes found differentially expressed than timepoints 3H and 9H: this is likely
due to biological or technical variances for specific genes at specific
timepoints.

@@

To leverage this temporal structure, Storey et al \citep{storey:significance} proposes to
model time-course micro-array with splines functions, and to use a log-ratio
likelihood test to detect differentially expressed genes. \texttt{moanin} extends
this idea by providing functions to perform time-course differential
expression analysis using \texttt{limma} contrasts.

@@

>>>>

\textbf{EAP: General comment. Your function names are incredibly long compared to standard functions. Also, will want to get rid of \texttt{moanin::} in front of functions. It's more precise, but an average user doesn't ever experience the \texttt{::} command; usually just package creaters. }

@@

The per-time point analysis is using classical differential expression approaches, and is often the approach advocated when dealing with small time-course datasets, where there are only a few time points \textbf{FIXME add citation to edgeR/limma vignette or DESEQ2}. For long time-course datasets, however, a separate test for each time-point results in creating many different tests, for example one for every time point, the results of which are difficult to integrate. We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.

@@

\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of these types of approaches, though our focus is on the global approach. In both situations, we first need to set up a object (a \texttt{splines\_model} object) to hold the meta data, as well as information for fitting the spline model.

@@

\textbf{EAP: General comment: you need a few more print() and head() of some objects to demonstrate what the output looks like}

@@

\texttt{moanin} provides a simple interface to perform a timepoint by timepoint differential
expression analysis. This is traditionally done by the user defining the comparisons (called \texttt{contrasts} in linear models). Under the hood, simply calls \texttt{limma} \citep{ritchie:limma} on the set of
contrasts provided. \textbf{EAP: Can we comment on using it for RNA-Seq??}

Here, we show an example where we define our contrasts to be the difference between the control mouse (``M'') and the
mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
(H5N1) (``VL'') for each time point, but the function works with any form contrasts. \textbf{FIXME: Reference to limma vignette for how to define the contrasts} \textbf{EAP: Could you make simple function that creates them automatically for two conditions -- i.e.~find union of time points for two conditions\ldots{} Then we could show an example with just a couple done manually, and then say we can also use a little helper function in the \texttt{moanin} that will do it for all time points}

@@

The distribution of number of genes found differentially expressed by
considering each time-point independantly highlights the challenges of such
approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
genes found differentially expressed than timepoints 3H and 9H: this is likely
due to biological or technical variances for specific genes at specific
timepoints. \textbf{EAP: We should add more to this}

@@

To leverage this temporal structure, Storey et al \citep{storey:significance} proposed to
model each gene in time-course micro-array with a splines function, and to use a log-ratio likelihood test to detect differentially expressed genes.

\texttt{moanin} extends
this idea by providing functionality to compare time course data between different treatment conditions, using a similar mechanism of contrasts -- only now the contrasts are differences between the estimated means.

\textbf{FIXME add in equations, like from the EPICON paper, to describe them mathematically}

\textbf{EAP: Do we need this call to limma by the user, or can these not be done internally by the \texttt{timecourse\_differential\_expression\_analysis} function? We can still preserve the option of user defining their own, but would be nicer to make this simpler for the user (e.g.~an if clause that tests whether input is matrix or character vector)}

@@

\textbf{EAP: \texttt{epicon} is going to be weird for general use}
